---
title: "Lesson 3: From LM to LMM"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Learning objectives

At the end of this lesson, students will ...

- Be able to fit a linear model with continuous and categorical predictor.
- Be able to fit a linear model with interaction terms.
- Understand the difference between a linear model and a linear mixed model.
- Be able to fit a mixed model with fixed effect and random intercept.
- Examine summary information from mixed model output, including ANOVA tables.

# Load some sample data

We are going to use some data from Ag Data Commons. I've simplified the data and added some random noise to it as well.

```{r}
library(tidyverse)
library(lmerTest)

barnyard_ghg <- read_csv('https://data.nal.usda.gov/system/files/BYD_GHG.csv')
```

Let's look at the data. Typing the name of the data frame gives you the first few rows.

```{r}
barnyard_ghg
```

Or you might prefer to use `str()`

```{r}
str(barnyard_ghg)
```

There is a column called `Surface` which contains character data. Look at the distribution of values in it with the `table()` function:

```{r}
table(barnyard_ghg$Surface)
```


# Model Syntax in R

The bare minimum we need is formula and data.

The formula consists of a left-hand side (LHS), a tilde `~`, and a right-hand side (RHS).

The left-hand side is the name of the response variable `y`. In this lesson we are only going to be focusing on models with a single response variable.

The right-hand side contains the predictor variables `x`. Effects are separated by a plus sign `+`. Interaction effects are denoted with a colon `:`. The shorthand `*` indicates all possible combinations of interactions for a set of variables. I do not recommend using this shorthand but you may see it. Another shorthand is `.` which means all variables in the dataset.

We will cover random effect syntax later. 

So a typical formula might be

`weight ~ height + sex + height:sex`

(In SAS this would be `model weight = height sex height*sex`). This means the expected value of weight is a linear combination of height (a continuous variable), sex (a binary variable), and the product or interaction of height and sex.

The `data` argument tells the model fitting function what data frame contains the variables in the model formula.

# Linear model with categorical predictor

This is a one-way ANOVA. We are looking at whether the mean CO2 flux from barnyard soil differs by surface type.

```{r}
fit1 <- lm(CO2_mgC ~ Surface, data = barnyard_ghg)
```

## Diagnostic plots

```{r}
plot(fit1)
```

What is wrong here?

The error isn't normally distributed. This is much better.

```{r}
fit2 <- lm(log(CO2_mgC) ~ Surface, data = barnyard_ghg)

plot(fit2)
```

Examine the output using the `summary()` function.

```{r}
summary(fit2)
```

You can see that there are three coefficients in the model: the intercept, sand, and soil.


We can include continuous effects, categorical effects, and interactions between any of them.

```{r}
fit_temp <- lm(log(CO2_mgC) ~ Temp, data = barnyard_ghg)
fit_surfacetemp <- lm(log(CO2_mgC) ~ Surface + Temp, data = barnyard_ghg)
fit_int <- lm(log(CO2_mgC) ~ Surface + Temp + Surface:Temp, data = barnyard_ghg)
fit_int_shortcut <- lm(log(CO2_mgC) ~ Surface*Temp, data = barnyard_ghg)
```

# Separate regressions for each sampling unit

People get hung up on checking relatively less important assumptions like normality of residuals. But a far more important assumption is that your data points are independent. If you fit a model to independent data points when it is truly non-independent, the results do not reflect reality. This is a much worse problem than the residuals not being normal.

In this example model fit, we have 1981 degrees of freedom. That is far too many. 

We can look in the data and see that samples were taken in a bunch of different barnyards. 

Okay, so what if there is a different relationship between CO2 flux and temperature in each barnyard? One way you might think about dealing with this is to fit a separate regression model to each barnyard.

I will demonstrate how you might do this -- note this code is a little bit more complex than what we'll be covering in this workshop so take this more as a demonstration. It is not something you would typically want to do.

We will `group_by()` and `nest()`.

```{r}
barnyard_ghg
```

