---
title: "Lesson 1: R basics"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Learning objectives

At the end of this lesson, students will ...

- Know what R is and what it can do.
- Use the R console to interactively issue R commands.
- Know what a data frame is.
- Know what R packages are and how to install and load them.
- Read a CSV file of data into the R workspace.
- Understand "tidy" data.
- Calculate summary statistics by group.
- Make a few basic plots.

# What is R?

# Using the R console

Here are some simple things you can do with R.

```{r}
2 + 3

x <- 2
y <- 3

x + y
x * 4

x <- x + 1
z <- x * 4
z

log(1000)

sin(pi)

my_name <- "Quentin"

paste("Hello,", my_name)
```

# What is a data frame?

We will not go into every possible R data type for this lesson. The one you really need to know about for model fitting is the data frame.

Load a package, read in an external file, plot the data, and fit a model.

```{r}
library(tidyverse)
library(lmerTest)

dat <- read_csv('testdata.csv')
```

```{r, fig.width = 4, fig.height = 4}
ggplot(dat, aes(x = x, y = y, color = g)) +
  geom_point()
```

```{r}
fit <- lmer(y ~ x + (1|g), data = dat)
```

```{r}
fixef(fit)
ranef(fit)
anova(fit)
```

