---
title: "Lesson 1: R basics"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Learning objectives

At the end of this lesson, students will ...

- Know what R is and what it can do.
- Use the R console to interactively issue R commands.
- Know what a data frame is.
- Know what R packages are and how to install and load them.

# Introduction to this course

- We're going to learn about mixed models in R.
- The course is divided up into lessons. For each lesson, I'm going to give you a worksheet. The worksheet will have a bunch of code in it, but parts of the code will be missing and replaced with a `...`. 
- Your job will be to fill in the `...` with the code. Don't worry, you will have access to the full code on the lesson webpage as well as being able to see it when I present it on the screen.
- I strongly recommend TYPING IN THE CODE yourself instead of copy+pasting it. Research shows that this helps you learn it better. Also, if you make a mistake and get an error or different output than you expected, you might learn something about how the code works (or doesn't work).

# What is R?

- A statistical programming language
- Users contribute packages
- Free and open-source

# What is RStudio?

RStudio is a tool to help you write and run code in R.

There are four panes that you see when you open RStudio:

- Console: Here you can enter R code line-by-line and run it interactively. The output of the code appears here.
- Scripts: This lets you type code into scripts (text files). You can run code from the script pane, or save it to run later.
- Environment: This shows what variables are currently defined in your R workspace, among other things. We won't worry about the other tabs at the moment.
- Files/plots/help: This pane has a few different tabs that you need to know about. The Files tab lets you navigate through directories on your system, the Plots tab shows any plots or images generated by your R code, the Packages tab shows you what packages are installed and lets you install other ones, and the Help tab shows documentation for functions and packages.

# What is R?

A language and an interpreter.

Everything in R is either a variable or a function.

(describe functions here)

# Using the R console

Here are some simple things you can do with R.

Things to note here: if you assign the result to a new variable, nothing will print to the console. But you can use the variable again later, as we do with `x` and `y` below.

```{r}
2 + 3

x <- 2
y <- 3

x + y
x * 4

x <- x + 1
z <- x * 4
z

log(1000)

sin(pi)

my_name <- "Quentin"

paste("Hello,", my_name)

?paste
```

# What kind of data can you work with in R?

You might have noticed the `[1]` before some of the output we made. Why is that `[1]` there? It means that the variable is a vector of length 1. We can make longer vectors as well.

This code also demonstrates how to use named arguments to a function.

```{r}
seq(from = 1, to = 100, by = 1)

1:100
```

Those are integer vectors. There are other data types, including character. We put character data in quotes.

This code also demonstrates how to index vectors, and the use of the `c()` function.

Single and double quotes are usually interchangeable.

```{r}
c('a', 'b', 'c', 'd', 'e', 'f', 'g')

letters[1:7]

letters[c(1, 18, 19)]
```

If you pass the wrong data type to a function, you usually get an error.

```{r}
log('hello')
```

If you try to combine numeric and character data in one vector, R will usually turn the whole thing into character. This is a common issue when loading data.

```{r}
c(100, 5.323, 'this is not a number', 12)
```


Factor is another important data type. It appears like a character but it can only contain predefined values. These values, called "levels," can be sorted in any order. This is useful for model fitting if you have a categorical predictor or response variable. If you specify the order of the levels, the first one is usually treated as the control or reference level in models. Factors are a little bit confusing but they are important to be aware of.

Here is an example factor.

```{r}
treatment <- factor(c('low', 'low', 'medium', 'medium', 'high', 'high'))

treatment
```

Re-sort the levels so that they are sorted in a logical order instead of alphabetical.

```{r}
treatment <- factor(treatment, levels = c('low', 'medium', 'high'))

treatment
```

Logical is another important data type. It can take two values, `TRUE` and `FALSE`. We usually get a logical vector when we do a comparison.

```{r}
x <- 1:5

x > 4

x <= 2

x == 3

x != 2

x > 1 & x < 5

x <= 1 | x >= 5
```


Missing data is coded as `NA`.

Some functions work on vectors and return one value for each element in the vector. Others return only one or a few values. Here are some examples of each.

# Common pitfalls

If you get an error or your code doesn't work, here are some things to check.

- Punctuation: close all parentheses, brackets, and quotation marks. (Notice parentheses can be used both for function arguments and to group expressions together.)

```{r}
(5+3))/2 # Nope

(5+3)/2 # Yep
```

- Spelling: are the functions and variables spelled correctly?

```{r}
my_variable <- 100000

myvariable
```

- Spaces: You can put spaces in between most things in R to make the code more readable, but never break up the name of a function or variable.

```{r}
some_numbers <- 1:5

( some_numbers + 3 ) ^ 2

(some_numbers+3)^2

(some numbers + 3)^2
```


- Case: R is CASE-SENSITIVE.

```{r}
sum(1:10)
Sum(1:10)
```

# How do I get help?

Google is your friend. Especially try to copy and paste your error message.

StackOverflow is your friend too (and stats.stackexchange.com if you have a question about stats that isn't specific to R programming). As a beginner, it is almost a certainty that you will find your question already answered on there. If you can't, read their guidelines on how to create a great reproducible example in R and post your question there!

# Statistical distributions

Talk about dnorm, pnorm, qnorm, and rnorm.

All distributions have 

- r: random numbers
- d: probability density (what is the y if you know an x)
- p: what is the cumulative probability at x
- q: quantile (what is the x if you know the cumulative probability or p)

# Console versus script editor

(describe how we can just type things into the editor, but we will use scripts to save and run again later)

We can run individual lines or selected blocks of code from the script editor by pressing `Ctrl+Enter` or `Cmd+Enter` while the cursor is on the line or some code is selected.

# Loading packages

Note we are going to use tidyverse packages.

# Hey! What about ... ?

- Functions
- Lists
- Flow control (if, else, for)

Those are really important things but we aren't going to cover them in this lesson. Functions and lists are really important if you start to do more complex analyses in R where you have to write your own code and can no longer use pre-packaged "off the shelf" functions to do everything you need to do. If, else, and for maybe aren't quite as crucial but still very important. I will not cover them today but I strongly encourage you to explore the R resources I've provided to learn more. And maybe I'll discuss them in a future workshop.

# Exercises

TBD