---
title: "Lesson 1: R basics"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE, error = TRUE)
```

# Learning objectives

At the end of this lesson, students will ...

- Know what R is and what it can do.
- Use the R console to interactively issue R commands.
- Know what a data frame is.
- Know what R packages are and how to install and load them.

# Introduction to this course

- We're going to learn about mixed models in R.
- The course is divided up into lessons. For each lesson, I'm going to give you a worksheet. The worksheet will have a bunch of code in it, but parts of the code will be missing and replaced with a `...`. 
- Your job will be to fill in the `...` with the code. Don't worry, you will have access to the full code on the lesson webpage as well as being able to see it when I present it on the screen.
- I strongly recommend TYPING IN THE CODE yourself instead of copy+pasting it. Research shows that this helps you learn it better. Also, if you make a mistake and get an error or different output than you expected, you might learn something about how the code works (or doesn't work).

# A practical toolkit

- This course is intended to teach some practical skills.
- It is NOT intended to teach elementary principles of statistics. There's no way to do that in two days!
- We may gloss over some technical details of the models. I'm happy to answer questions about them later.

# What is R?

- A statistical programming language
- Users contribute packages
- Free and open-source

# What is RStudio?

RStudio is a tool to help you write and run code in R.

There are four panes that you see when you open RStudio:

- Console: Here you can enter R code line-by-line and run it interactively. The output of the code appears here.
- Scripts: This lets you type code into scripts (text files). You can run code from the script pane, or save it to run later.
- Environment: This shows what variables are currently defined in your R workspace, among other things. We won't worry about the other tabs at the moment.
- Files/plots/help: This pane has a few different tabs that you need to know about. The Files tab lets you navigate through directories on your system, the Plots tab shows any plots or images generated by your R code, the Packages tab shows you what packages are installed and lets you install other ones, and the Help tab shows documentation for functions and packages.

# The basic moving parts of R

R is both a language and an interpreter.

Everything in R is either a variable or a function.

- **variable**: a structure that holds data. Examples:
  + a vector of integers `c(1, 2, 3)`
  + a character string `"USDA"`
  + a data frame with 1000 rows and 10 columns... all these things are variables
  
- **function**: something that takes arguments as input, does something, and returns output. Examples:
  + `log(10)`: The function `log()` takes `10` as input and returns output.
  + `c(1, 5, 6)`: The function `c()` takes three numbers as input and returns a vector as output.
  + `read.csv('myfile.csv')`: The function `read.csv()` takes a character string with a the file name of a CSV file as input, reads that CSV file, and returns a data frame as output.

# Using the R console

Here are some simple things you can do with R.

Things to note here: 

- if you assign the result to a new variable, nothing will print to the console. But you can use the variable again later, as we do with `x` and `y` below.
- any line preceded by `#` is a comment and will not be evaluated.

```{r}
2 + 3

x <- 2
y <- 3

x + y
x * 4

x <- x + 1
z <- x * 4
z

log(1000)

sin(pi)

# This is a comment.

my_name <- "Quentin"

paste("Hello,", my_name)
```

```{r, eval = FALSE}
?paste
```

# What kind of data can you work with in R?

You might have noticed the `[1]` before some of the output we made. Why is that `[1]` there? It means that the variable is a vector of length 1. We can make longer vectors as well.

This code also demonstrates how to use named arguments to a function.

```{r}
seq(from = 1, to = 100, by = 1)

1:100
```

You can see that the output provides a few indexes along the left side.

Those are integer vectors. There are other data types, including character. We put character data in quotes.

This code also demonstrates how to index vectors, and the use of the `c()` function.

Single and double quotes are usually interchangeable.

```{r}
c('a', 'b', 'c', 'd', 'e', 'f', 'g')

letters[1:7]

letters[c(1, 18, 19)]
```

If you pass the wrong data type to a function, you usually get an error.

```{r}
log('hello')
```

If you try to combine numeric and character data in one vector, R will usually interpret the whole thing as character data. This is a common issue when loading data.

```{r}
c(100, 5.323, 'missing value', 12)
```


Factor is another important data type. It appears like a character but it can only contain predefined values. These values, called "levels," can be sorted in any order. This is useful for model fitting if you have a categorical predictor or response variable. If you specify the order of the levels, the first one is usually treated as the control or reference level in models. Factors are a little bit confusing but they are important to be aware of.

Here is an example factor.

```{r}
treatment <- factor(c('low', 'low', 'medium', 'medium', 'high', 'high'))

treatment
```

Re-sort the levels so that they are sorted in a logical order instead of alphabetical.

```{r}
treatment <- factor(treatment, levels = c('low', 'medium', 'high'))

treatment
```

Logical is another important data type. It can take two values, `TRUE` and `FALSE`. We usually get a logical vector when we do a comparison.

```{r}
x <- 1:5

x > 4

x <= 2

x == 3

x != 2

x > 1 & x < 5

x <= 1 | x >= 5
```


Missing data is coded as `NA`.

Some functions work on vectors and return one value for each element in the vector. Others return only one or a few values. Here are some examples. The function `exp()` takes a vector as input and returns a vector of the same length, the exponential of each element in the vector. Notice I use `head()` to avoid printing all 1000 values. The functions `length()`, `mean()`, `median()`, and `sd()` take a vector as input and return a single value. The function `range()` returns a vector of two values, the minimum and maximum of the vector. The function `quantile()` is a little bit more complicated. It takes two vectors as input. The second vector, `probs`, contains the probabilities we want to calculate the quantiles for. The function returns a vector with the same length as `probs` containing the percentiles.

> PROTIP: We use `set.seed()` to ensure the code produces the same result each time we generate random numbers.

```{r}
set.seed(123)

random_numbers <- rnorm(n = 1000, mean = 0, sd = 1)

head(exp(random_numbers))

length(random_numbers)

mean(random_numbers)

median(random_numbers)

sd(random_numbers)

range(random_numbers)

quantile(random_numbers, probs = c(0.025, 0.5, 0.975))
```

# Statistical distributions

In the code just above, we saw `rnorm()`, which generates random draws from a normal distribution. R has a lot of built-in statistical distributions. All of them have four functions beginning with `r`, `d`, `p`, and `q`, and followed by the (abbreviated) name of the distribution. 

The four functions are:

- `r`: random draws from the distribution
- `d`: probability density function (what is the y-value of the function given x?)
- `p`: cumulative density function: (what is the cumulative probability given x?)
- `q`: quantile (what is the x-value given the cumulative probability?); `q` is the inverse of `p`.

For example, the functions for the normal distribution are `rnorm()`, `dnorm()`, `pnorm()`, and `qnorm()`. All of these default to the standard normal distribution with `mean = 0` and `sd = 1`, but you can change those parameters by modifying the `mean` and `sd` arguments.

These figures illustrate what `dnorm()`, `pnorm()`, and `qnorm()` do.

```{r, echo = FALSE, fig.show = 'hold', out.width = '50%', fig.width = 4.5, fig.height = 4.5}
library(ggplot2)
windowsFonts(`consolas` = windowsFont("Consolas"))

ggplot() +
  stat_function(fun = dnorm, xlim = c(-4, 4)) +
  scale_x_continuous(breaks = -3:3) +
  scale_y_continuous(name = 'y', expand = expansion(mult = c(0, 0.1))) +
  coord_cartesian(xlim = c(-3, 3)) +
  geom_point(aes(x = 0, y = dnorm(0)), color = 'indianred', size = 3) +
  geom_segment(aes(x = 0, xend = 0, y = 0, yend = dnorm(0)), color = 'indianred', linetype = 'dashed') +
  geom_segment(aes(x = -Inf, xend = 0, y = dnorm(0), yend = dnorm(0)), color = 'indianred', linetype = 'dashed') +
  geom_text(aes(x = 0, y = 0.4), label = 'dnorm(0) = 0.399', family = 'consolas', vjust = -1, size = 6) +
  theme_classic()
  

ggplot() +
  stat_function(fun = dnorm, xlim = c(-4, 4)) +
  scale_x_continuous(breaks = -3:3) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  coord_cartesian(xlim = c(-3, 3)) +
  stat_function(geom = 'area', xlim = c(-4, 0), fun = dnorm, fill = 'slateblue') +
  geom_text(aes(x = 0, y = 0.4), label = 'pnorm(0) = 0.5', family = 'consolas', vjust = -1, size = 6) +
  geom_text(aes(x = 0.1, y = 0.05), label = 'qnorm(0.5) = 0', family = 'consolas', hjust = 0, size = 6) +
  theme_classic()
```


Other distributions you might work with:

- Binomial (`rbinom()`, `dbinom()`, `pbinom()`, `qbinom()`)
- Uniform (`runif()`, `dunif()`, `punif()`, `qunif()`)
- Student's *t* (`rt()`, `dt()`, `pt()`, `qt()`)
- The list goes on ...

Type `?Distributions` in your console to see help documentation about all the built-in distributions. Many R packages have implemented other less common distributions.

# Common pitfalls

If you get an error or your code doesn't work, here are some things to check.

- Punctuation: close all parentheses, brackets, and quotation marks. (Notice parentheses can be used both for function arguments and to group expressions together.)

```{r}
(5+3))/2 # Nope

(5+3)/2 # Yep
```

- Spelling: are the functions and variables spelled correctly?

```{r}
my_variable <- 100000

myvariable
```

- Spaces: You can put spaces in between most things in R to make the code more readable, but never break up the name of a function or variable.

```{r}
some_numbers <- 1:5

( some_numbers + 3 ) ^ 2

(some_numbers+3)^2

(some numbers + 3)^2
```


- Case: R is CASE-SENSITIVE.

```{r}
sum(1:10)
Sum(1:10)
```

# How do I get help?

Google is your friend. Especially try to copy and paste your error message.

StackOverflow is your friend too (and stats.stackexchange.com if you have a question about stats that isn't specific to R programming). As a beginner, it is almost a certainty that you will find your question already answered on there. If you can't, read their guidelines on how to create a great reproducible example in R and post your question there!

# Console versus script editor

(describe how we can just type things into the editor, but we will use scripts to save and run again later)

We can run individual lines or selected blocks of code from the script editor by pressing `Ctrl+Enter` or `Cmd+Enter` while the cursor is on the line or some code is selected.

# Loading packages

Note we are going to use tidyverse packages.

# Hey! What about ... ?

- Functions
- Lists
- Flow control (if, else, for)

Those are really important things but we aren't going to cover them in this lesson. Functions and lists are really important if you start to do more complex analyses in R where you have to write your own code and can no longer use pre-packaged "off the shelf" functions to do everything you need to do. If, else, and for maybe aren't quite as crucial but still very important. I will not cover them today but I strongly encourage you to explore the R resources I've provided to learn more. And maybe I'll discuss them in a future workshop.

# Exercises

- Use R to find the arc-sine of the square root of 0.5. (Hint: you may need to use help documentation or Google to figure out what function or functions to use).
- Use R to find the sum of all positive integers from 1 to 12345.
