---
title: "Lesson 3: From LM to LMM"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = TRUE)
```

# Learning objectives

At the end of this lesson, students will ...

- Be able to fit a linear model with continuous and categorical predictor.
- Be able to fit a linear model with interaction terms.
- Understand the difference between a linear model and a linear mixed model.
- Be able to fit a mixed model with fixed effect and random intercept.
- Examine summary information from mixed model output, including ANOVA tables.

# Load some sample data

We are going to use some data from the **agridat** package. You can either load the example data from the package directly using `data()`, or read it in from a CSV in the `datasets` folder. Here are both ways of doing it.

```{r}
library(tidyverse)
library(lmerTest)

barnyard_ghg <- read_csv('https://data.nal.usda.gov/system/files/BYD_GHG.csv') %>%
  filter(Days_elapsed == 0, !is.na(CO2_mgC)) %>%
  select(Barnyard, Surface, Temp, CO2_mgC, NH3_mg, N2O_mg)
```

Let's look at the data. Typing the name of the data frame gives you the first few rows.

```{r}
barnyard_ghg
```

Or you might prefer to use `str()` or `glimpse()`.

```{r}
glimpse(barnyard_ghg)
```

There is a numeric column to identify which barnyard the measurement comes from, a character column for what the surface type is, and numeric columns for temperature, CO2 emissions, NH3 emissions, and N2O emissions. Some of the values are missing (`NA`).

There is a column called `Surface` which contains character data. Look at the distribution of values in it with the `table()` function:

```{r}
table(barnyard_ghg$Surface)
```


# Model Syntax in R

The bare minimum we need is formula and data.

The formula consists of a left-hand side (LHS), a tilde `~`, and a right-hand side (RHS).

The left-hand side is the name of the response variable `y`. In this lesson we are only going to be focusing on models with a single response variable.

The right-hand side contains the predictor variables `x`. Effects are separated by a plus sign `+`. Interaction effects are denoted with a colon `:`. The shorthand `*` indicates all possible combinations of interactions for a set of variables. I do not recommend using this shorthand but you may see it. Another shorthand is `.` which means all variables in the dataset.

We will cover random effect syntax later. 

So a typical formula might be

`weight ~ height + sex + height:sex`

(In SAS this would be `model weight = height sex height*sex`). This means the expected value of weight is a linear combination of height (a continuous variable), sex (a binary variable), and the product or interaction of height and sex.

The `data` argument tells the model fitting function what data frame contains the variables in the model formula.

# Linear model with categorical predictor

This is a one-way ANOVA. We are looking at whether the mean CO2 flux from barnyard soil differs by surface type.

```{r}
fit_oneway <- lm(CO2_mgC ~ Surface, data = barnyard_ghg)
```

## Diagnostic plots

```{r}
plot(fit_oneway)
```

Press `<Enter>` to bring up each plot in order.

These diagnostic plots look reasonable. In particular, the first plot shows that the distributions of residuals are about the same for the three surface types, indicating there is no relationship between residual and fitted values (the assumption of homogeneous variance is met). The second plot shows that quantiles of the residuals fall very close to a straight line relative to the quantiles of a normal distribution. This indicates the residuals are roughly normally distributed. 

Examine the output using the `summary()` function.

```{r}
summary(fit_oneway)
```

You can see that there are three coefficients in the model: the intercept, sand, and soil. Even though there are three surfaces (bark, sand, and soil), only two of them (sand and soil) have their own coefficient. This is because we have an intercept already, and the mean value of one of the three surfaces is set as the intercept. Then we only need two additional coefficients to define the means for the other two levels. "Bark" is set as the reference level because it is the first in alphabetical order. This is the default, which you can change by changing the order of factor levels.

What the summary tells us is that the expected mean value of CO2 emissions from bark is about 475.6 mg C. The coefficients on sand and soil are negative, so the emissions are lower from both of those surfaces, relative to bark. For example, the expected value of emissions from sand is 476.6 - 169.2 which is ~306.4 mg C. We will look at ways to compare these means later.

We also see that there are built-in t-tests for each coefficient. The t-test for the intercept is not really meaningful because it is comparing the intercept against a value of zero. Of course all the CO2 fluxes are well above zero so it's not surprising to get a high t-statistic and low p-value here. However the other two t-tests are testing whether the difference between two surfaces (sand and bark, and soil and bark, respectively), is not zero. So those are meaningful. They show that both the other two surfaces are different from bark. However these don't account for multiple comparisons or anything, and there is no t-test showing whether sand and soil differ from one another, so they aren't the best ones to report. We'll revisit that tomorrow!

# ANOVA

We can get an ANOVA table for the linear model using the function `anova()`.

```{r}
anova(fit_oneway)
```

# Plot of the model results

Here's a plot of the model results. For now I will not explain how it was made. You will learn about that tomorrow.

```{r}
# FIXME include this here.
```


# Different effects

The `barnyard_ghg` dataset also has a variable `Temp` indicating the temperature at the time the gas flux measurement was taken. It is a continuous variable.

We can include continuous effects, categorical effects, and interactions between any of them. Notice that I've written 

```{r}
fit_temp <- lm(CO2_mgC ~ Temp, data = barnyard_ghg)
fit_surfacetemp <- lm(CO2_mgC ~ Surface + Temp, data = barnyard_ghg)
fit_int <- lm(CO2_mgC ~ Surface + Temp + Surface:Temp, data = barnyard_ghg)
fit_int_shortcut <- lm(CO2_mgC ~ Surface*Temp, data = barnyard_ghg)
```

# Separate regressions for each sampling unit

People get hung up on checking relatively less important assumptions like normality of residuals. But a far more important assumption is that your data points are independent. If you fit a model to independent data points when it is truly non-independent, the results do not reflect reality. This is a much worse problem than the residuals not being normal.

In this example model fit, we have 106 degrees of freedom. That is far too many. 

We can look in the data and see that samples were taken in a bunch of different barnyards. 

```{r}
table(barnyard_ghg$Barnyard)
```

There are nine barnyards, each with about a dozen measurements. The measurements that come from the same barnyard are not independent because they have a common cause: whatever properties the soil at barnyard 1 has that influence CO2 emissions are affecting all the CO2 measurements from barnyard 1 in the same way. Thus, 106 degrees of freedom is probably too many. But if we just average all the measurements from each combination of barnyard and surface type into a single measurement, we will only have 6 degrees of freedom in the model. A more appropriate value would be somewhere in between -- the measurements from the same barnyard are *partially*, but not completely, dependent on one another.

INSERT MORE TEXT HEREREERERE

Okay, so what if there is a different relationship between CO2 flux and temperature in each barnyard? One way you might think about dealing with this is to fit a separate regression model to each barnyard.

I will demonstrate how you might do this -- note this code is a little bit more complex than what we'll be covering in this workshop so take this more as a demonstration. It is not something you would typically want to do.

We will `group_by()` and `nest()`.

```{r}
barnyard_separate_models <- barnyard_ghg %>%
  group_by(Barnyard) %>%
  nest %>%
  mutate(model = map(data, ~ lm(CO2_mgC ~ Surface, data = .)))
#FIXME put this in there
```

# Our first mixed model

It might seem like we have dealt with the issue that the different barnyards might have different relationships between surface type and CO2 emissions. But we went a little bit too far. We really want to know how surface type affects CO2 emissions in general, not in one particular barnyard ... the goal of our scientific inference is usually to generalize what we find in a particular study to a larger population of barnyards (only a few of which we actually visited to get CO2 measurements in our study). Doing separate regressions for each barnyard doesn't quite get us there because it doesn't give us *an estimate of the overall relationship between surface and CO2 emissions in barnyards, while at the same time accounting for variation due to the unknown factors that may be different from barnyard to barnyard*. 

How can we solve that problem? You probably already guessed it ... a mixed model! I'm calling these mixed models but they go by a lot of other names.

(dumb picture courtesy of Chelsea Parlett-Pelleriti on Twitter)

Mixed models do what is called "partial pooling." 

(insert some other description of mixed models)

Here's our first mixed model. We are now using the `lmer()` function from the **lme4** package instead of `lm()` from base R.

```{r}
fit_mm <- lmer(CO2_mgC ~ Temp, data = barnyard_ghg)
```

# Transformation of response variables

- Log transformation

# ANOVA

- How to get the F-tests

# Hey! What about ...

- These are only a little bit of what you can do. I don't have time to cover all GLMM types in this workshop.
- glmer() doesn't have all the possibilities
- Other common ones are beta-regression, where the response variable is a probability or proportion bounded between 0 and 1
- Cumulative logistic mixed model is another common case, where the response variable is a categorical outcome with more than two possibilities
  + For example a choice trial where insects are released into an olfactometer and may go down one of four arms where different odor attractants are found.
  
- The Bayesian modeling packages **brms** and **rstanarm** are also great alternatives that I would recommend looking into. I'm going to do a workshop on **brms** soon.

# What have we done so far

You've learned the basics of R. You've learned how to work with data frames in R. You've fit a linear model, a linear mixed model, and a generalized linear mixed model in R. Impressive!

# What we will do tomorrow

Tomorrow, we will do a lot more fun stuff with mixed models. We will cover ...

- Random slope models
- How to compare means and test specific hypotheses with contrasts
- How to make some basic plots and tables to present the results of your model

Excited to see you all then!!!