# ggplot2 section removed from lesson two

# Make a basic plot

We will use the **ggplot2** package for plotting. In this workshop I will not explain how these plots work in great detail, but I plan to offer workshops on data visualization in the future.

As a brief intro, we can "map" different columns of the data frame to different components of the plot. This is known as "aesthetic mapping" hence the function name `aes()`. For example, in this fake data frame, we map the `x` column to the x-axis aesthetic, the `y` column to the y-axis aesthetic, and the `g` column, which consists of letters, to the color aesthetic.

Then you use the plus sign `+` to add "geoms" to the plot which correspond to different types of data visualizations you are probably familiar with. For instance `geom_point()` is a scatterplot.

(insert cheat sheet)

```{r, fig.width = 4, fig.height = 4}
ggplot(dat, aes(x = x, y = y, color = g, fill = g)) +
  geom_point()
```

As you can see, different types of plots can be made by changing the geom, with the variable mappings unchanged, such as a line graph or a stacked bar plot. The data are the same.

```{r, fig.width = 4, fig.height = 4}
ggplot(dat, aes(x = x, y = y, color = g, fill = g)) +
  geom_line()

ggplot(dat, aes(x = x, y = y, color = g, fill = g)) +
  geom_col()
```

Additional arguments can be added to change the appearance of the plot, alter the geoms, or add other graphical elements that are not mapped from the data frame. For instance we can change our stacked bar plot to a grouped bar plot by an argument to `geom_col()`. Then we can change the theme by adding `theme_bw()`, and also add a horizontal dotted line on the plot by adding another geom.

```{r}
barplot <- ggplot(dat, aes(x = x, y = y, color = g, fill = g)) +
  geom_col(position = 'dodge')

barplot

barplot + theme_bw()

barplot + theme_bw() + geom_hline(yintercept = 15, linetype = 'dotted')
```